Linking watershed nutrient loading to estuary water quality with semiparametric models

### 2023-08-07 Schramm-2023-06-PeerJ/Schramm-2023-08-PeerJ.tex
start working on revisions

\usepackage{csquotes}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\begin{abstract}
Evaluating estuary water quality responses to reductions (or increases)
in nutrient loading attributed to on the ground management actions can
be challenging due to the strong influence of environmental drivers on
nutrient loads and non-linear relationships. This study applied
semiparametric approaches to calculate watershed nutrient loads and
assess responses in estuary water quality to seasonally-adjusted
freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas.
Lavaca Bay is a secondary embayment on the Texas coast displaying early
potential for eutrophication and water quality degradation. Use of
flow-adjusted nutrient loads allowed the study to evaluate the response
in water quality to changes in nutrient loads driven by non-natural
variation. Cross-validation indicated that, despite data constraints,
semiparametric models performed well at nutrient load prediction. Based
on these models, delivered annual nutrient loads varied substantially
from year to year. In contrast, minimal changes in flow-normalized loads
indicate that nutrient loadings were driven by natural variation in
precipitation and runoff as opposed to changes in management of nonpoint
sources. Models indicated no evidence of long-term changes in dissolved
oxygen or chlorophyll-\emph{a} within Lavaca Bay. However, site specific
long-term increases in both organic and inorganic nitrogen are
concerning for their potential to fuel eutrophication. Further analysis
found freshwater inflow had strong influences on nutrient and
chlorophyll-\emph{a} concentrations but there was no evidence that
changes in watershed nutrient loading explained additional variation in
dissolved oxygen or chlorophyll-\emph{a} concentrations. In addition to
providing a baseline assessment of watershed nutrient loading and water
quality responses in the Lavaca Bay watershed, this study provides
methodological support for the use of semiparametric models in load
regression models and estuary assessments.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

There are ongoing efforts between local, state, and federal agencies to
address water quality impairments in the freshwater portions of the
Lavaca Bay watershed
\autocite{jainTechnicalSupportDocument2021,schrammLavacaRiverWatershed2018,bertholdDirectMailingEducation2021}.
However, on a statewide scale, these approaches have shown limited
success and emphasize a need for improved methods to assess and link
management actions with downstream water quality
\autocite{schrammTotalMaximumDaily2022}.\\
Such methods could help identify and replicate effective water quality
management actions across the state . The identification and
communication of changes and trends in water quality is complicated by
the fact that trends are often non-linear and confounded by
precipitation and runoff that hinder traditional analysis
\autocite{wazniakLinkingWaterQuality2007,lloydMethodsDetectingChange2014}.
The development and application of flexible statistical methods such as
Weighted Regressions on Time, Discharge and Season
\autocite[WRTDS,][]{hirschWeightedRegressionsTime2010} and Generalized
Additive Models \autocite[GAMs,][]{woodFastStableRestricted2011} have
provided effective tools for researchers to quantify and communicate
non-linear changes in river and estuary pollutant loadings.

@@

While WRTDS is a statistical approach developed specifically for water
quality applications, GAMs are a broadly applicable statistical method.
GAMs are a semiparametric extension of generalized linear models where
the linear predictor is represented as the sum of multiple unknown
smooth functions and parametric linear predictors
\autocite{woodFastStableRestricted2011}. Although the underlying
parameter estimation procedure of GAMs are substantially different than
WRTDS, both the functional form and results have been demonstrated to be
similar when assessing nutrient concentration trends
\autocite{beckNumericalQualitativeContrasts2017}. Water quality
applications of GAMs have included river and catchment nutrient
concentration and load estimation
\autocite{wangLoadEstimationUncertainties2011,kroonRiverLoadsSuspended2012,kuhnertQuantifyingTotalSuspended2012,robson_prediction_2015-1,hagemannEstimatingNutrientOrganic2016,mcdowell_implications_2021,biagi_novel_2022},
and assessment of temporal trends of nutrients
\autocite{beckNumericalQualitativeContrasts2017,murphyGeneralizedAdditiveModel2019},
phytoplankton \autocite{bergbuschUnexpectedShiftPhytoplankton2021}, and
cyanobacteria \autocite{hayesEffectsLakeWarming2020}. Recently GAMs have
also been used to link water quality responses in receiving water bodies
to changes in nonpoint source nutrient inputs
\autocite{murphyNutrientImprovementsChesapeake2022}. For a substantial
discussion on the differences (and similarities) between GAMs and WRTDS
for water quality applications readers are referred to
\textcite{beckNumericalQualitativeContrasts2017}.

@@

Lavaca Bay is a 190 km\textsuperscript{2} estuary with the majority of
freshwater inflow provided by the Lavaca and Navidad River systems
(Fig.~\ref{fig:fig1}). The Garcitas-Arenosa, Placedo Creek, and Cox Bay
subwatersheds provide additional freshwater inflows. The entire
watershed area is 8,149 km\textsuperscript{2} and primarily rural.
Watershed land cover and land use is 50\% grazed pasture and rangeland,
20\% cultivated cropland (primarily rows crops such as corn, cotton, and
sorghum), and 5\% suburban/urban. Pasture and rangeland is concentrated
in the Lavaca River watershed, while cultivated crops are generally
located along the eastern tributaries of the Navidad river. The Lavaca
and Navidad River watersheds are a combined 5,966 km\textsuperscript{2},
or approximately 73\% of the entire Lavaca Bay watershed area. Discharge
from the Navidad River is regulated by Lake Texana which has been in
operation since 1980. Lake Texana provides 0.210 km\textsuperscript{3}
of water storage and discharges into the tidal section of the Navidad
River which ultimately joins the tidal section of the Lavaca River 15 km
upstream of the confluence with the Lavaca Bay.

@@

{\centering \includegraphics[width=5.2in,]{Schramm-2023-06-PeerJ_files/figure-latex/fig1}

@@

\caption{Map of Lavaca Bay and the contribution watershed. The freshwater sites are the most downstream freshwater stream locations with water quality and streamflow data used for nutrient load models. Water quality concentration data at the three Lavaca Bay sites were used to assess relationships between freshwater flows, loads and estuary water quality.}\label{fig:fig1}
\end{figure}

Daily discharges for the Lavaca River (USGS-08164000,
Fig.~\ref{fig:fig1}) were obtained from the United States Geologic
Survey (USGS) National Water Information System using the
\emph{dataRetrieval} R package
\autocite{deciccoDataRetrievalPackagesDiscovering2022}. Gaged daily
discharges from the outlet of Lake Texana on the Navidad River
(USGS-0816425) were provided by the Texas Water Development Board (TWDB)
(April 21, 2022 email from R. Neupane, TWDB).

@@

\caption{\label{tab:table1}Summary of gauged streamflow and freshwater water quality samples between January 1, 2000 and December 31, 2020.}
\centering
\begin{tabular}[t]{lllll}
\toprule
Station ID & Parameter & Mean & SD & N\\
\midrule
 & TP (mg/L) & 0.21 & 0.09 & 80\\

 & NO\textsubscript{3} (mg/L) & 0.18 & 0.24 & 74\\

\multirow{-3}{*}{\raggedright\arraybackslash USGS-08164000} & Mean Daily Streamflow (cfs) & 332.78 & 1667.47 & 7671\\
\cmidrule{1-5}
 & TP (mg/L) & 0.20 & 0.08 & 81\\

 & NO\textsubscript{3} (mg/L) & 0.29 & 0.26 & 62\\

\multirow{-3}{*}{\raggedright\arraybackslash USGS-08164525} & Mean Daily Streamflow (cfs) & 666.14 & 2957.79 & 7671\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:table2}Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.}
\centering
\begin{tabular}[t]{lllll}
\toprule
Station ID & Parameter & Mean & SD & N\\
\midrule
 & TP (mg/L) & 0.11 & 0.05 & 47\\

 & NO\textsubscript{x} (mg/L) & 0.07 & 0.15 & 51\\

 & TKN (mg/L) & 0.94 & 0.49 & 45\\

 & Chlorophyll-\textit{a} (ug/L) & 9.43 & 5.31 & 47\\

\multirow{-5}{*}{\raggedright\arraybackslash TCEQ-13383} & DO (mg/L) & 7.22 & 1.35 & 55\\
\cmidrule{1-5}
 & TP (mg/L) & 0.08 & 0.03 & 51\\

 & NO\textsubscript{x} (mg/L) & 0.06 & 0.08 & 52\\

 & TKN (mg/L) & 0.76 & 0.40 & 48\\

 & Chlorophyll-\textit{a} (ug/L) & 8.22 & 6.44 & 46\\

\multirow{-5}{*}{\raggedright\arraybackslash TCEQ-13384} & DO (mg/L) & 7.51 & 1.32 & 54\\
\cmidrule{1-5}
 & TP (mg/L) & 0.13 & 0.06 & 50\\

 & NO\textsubscript{x} (mg/L) & 0.09 & 0.13 & 53\\

 & TKN (mg/L) & 0.94 & 0.37 & 49\\

 & Chlorophyll-\textit{a} (ug/L) & 9.67 & 5.33 & 49\\

\multirow{-5}{*}{\raggedright\arraybackslash TCEQ-13563} & DO (mg/L) & 7.91 & 1.34 & 56\\
\bottomrule
\end{tabular}
\end{table}

@@

Estimates of NO\textsubscript{3} and TP loads at the Lavaca River
(USGS-08164000) and the outlet of Lake Texana on the Navidad River
(USGS-08164525) were developed using GAMs relating nutrient
concentration to river discharge, season, and time. Separate models were
fit at each station for each parameter and used to predict nutrient
concentrations for each day in the study period. GAMs were fit using the
\emph{mgcv} package in R which makes available multiple types of smooth
functions with automatic smoothness selection
\autocite{woodFastStableRestricted2011}. The general form of the model
related NO\textsubscript{3} or TP concentration to a long term tend,
season, streamflow, and two different antecedent discharge terms, shown
in Eq. \ref{eq1}:

@@

The calculation of model terms for the Lake Texana site were modified
because daily loads are not a function of natural stream flow processes
alone, but of dam releases and nutrient concentrations at the discharge
point of the lake. \emph{Q}, \emph{ma}, and \emph{fa} terms were
calculated based on total gaged inflow from the 4 major tributaries to
the lake. Thin-plate regression splines were used for \emph{ddate},
\emph{log1p(Q)}, \emph{fa}, and \emph{ma}. A cyclic cubic regression
spline was used for \emph{yday} to ensure the ends of the spline match
(day 1 and day 366 are expected to match). First order penalties were
applied to the smooths of flow-based variables which penalize departures
from a flat function to help constrain extrapolations for high flow
measurements.

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig2}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig3}

@@

{\centering \includegraphics[width=0.5\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig4}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig5}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig6}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig7}

@@

Some prior studies have generated estimates of mean annual TP yields in
the Lavaca River watershed
\autocites[Table~\ref{tab:table5},][]{dunnTrendsNutrientInflows1996,rebichSourcesDeliveryNutrients2011,omaniEstimationSedimentNutrient2014,wise_spatially_2019}.
Although these studies differ in time periods and methodologies, they
provide a sanity check for the reasonableness of the annual estimates
generated in the current study. In a regional assessment of nutrient
loading in river's along the Gulf of Mexico,
\textcite{dunnTrendsNutrientInflows1996} used the LOADEST model to
develop an estimated mean annual yield of 28.9 kg/km\textsuperscript{2}.
LOADEST is a multiple linear regression model that fits log transformed
pollutant concentrations to long term, seasonal, and flow based
predictors and includes methods for bias correction when exponentiating
the response variable. While the annual yield estimated in
\textcite{dunnTrendsNutrientInflows1996} is lower than the estimate of
35.2 kg/km\textsuperscript{2} in this study, it is important to note
that both hydrology and land use was different for a proportion of the
study period. For almost the first ten years of the study period in
\textcite{dunnTrendsNutrientInflows1996}, the Navidad River was still
undammed. More recently the eastern portion of the watershed has seen
large conversions of rice fields to more traditional row crop production
\autocite{kulatHolisticWaterEnergyFoodNexus2019}. The apparent increase
in mean TP yield might be suggestive of wetter recent conditions. To
evaluate long-term trends in discharge, we fit a GAM relating
log-transformed daily discharges on the Lavaca and Navidad Rivers to
season and time (Fig.~\ref{fig:fig8}). The long-term trends in discharge
indicate watershed discharges were at or above average from 1972 though
the early and mid-1980s. In comparison watershed discharges since the
mid-2000s are at or below average. The trend in mean streamflow alone is
counter to expectations based on changes in nutrient yeild estimates.
Studies of long-term precipitation patterns indicate both the total
amount \autocite{dixonTrendDetectionTexas2011} and variability of
rainfall
\autocite{mishraChangesExtremePrecipitation2010,fagnantCharacterizingSpatiotemporalTrends2020}
has increased in the region which provide plausible reasons for
increased runoff driven nutrient loading. Further assessments of the
frequency and occurrence of periodic high flow events that mobilize
sediments and are responsible for the bulk of loading volume would
provide further insight to this discrepancy.

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-06-PeerJ_files/figure-latex/fig8}

@@

\caption{Measured daily discharges (log-transformed) and smoothed long-term trends for the Lavaca and Navidad Rivers form 1972 though 2001.}\label{fig:fig8}
\end{figure}

\textcite{rebichSourcesDeliveryNutrients2011} and
\textcite{wise_spatially_2019} used SPARROW to provide a more recent
assessment of regional catchment based loadings to the Gulf of Mexico
(Table~\ref{tab:table5}). SPARROW is a hybrid statistical-process model
with the underlying nutrient load estimation methods based on the
previously described LOADEST
\autocite{schwarzSPARROWSurfaceWaterQuality2006}. The functional form of
the LOADEST regression model is similar to the terms applied in the GAMs
used in the current study. The only study to apply a mechanistic
watershed model (SWAT) to estimate nutrient loadings in the Lavaca River
watershed \textcite{omaniEstimationSedimentNutrient2014} developed
estimated yields (42 kg/km\textsuperscript{2}) similar to the two
SPARROW models. The base periods used by
\textcite{rebichSourcesDeliveryNutrients2011},
\textcite{omaniEstimationSedimentNutrient2014} and
\textcite{wise_spatially_2019} differ from the current study, making
direct comparisons difficult. It is probable that the lower than average
discharges observed from 2010 through 2021 (Fig.~\ref{fig:fig8}) bias
our estimates downward compared to other studies which included higher
than average streamflow periods. Overall, the ranges of estimated yields
among different studies along with the apparent large variability in
streamflow driven loadings (Fig.~\ref{fig:fig3}, Fig. \ref{fig:fig4})
suggest that the current estimates of TP loading are reasonable.

@@

There was no evidence that adjusted freshwater inflow and nutrient loads
had effects on DO concentration in Lavaca Bay. The seasonality term in
the temporal GAM models explained a substantial amount of DO variation
at all of the sites. Responses of estuary metabolic processes and
resulting DO concentrations can be quite complicated and often locally
specific \autocite{caffreyFactorsControllingNet2004}. While the lack of
total nitrogen or TKN loading data hinders interpretation, the large
seasonal effect on DO concentration indicates physical factors (such as
temperature, wind, and turbidity) play an important role and should be
included in future models. Prior work suggests that Lavaca Bay may not
be limited by nutrients alone, with high turbidity or nutrient
processing in upper portions of the Bay or intertidal river limiting
production \autocite{russell_effect_2006}. Finally, it is reasonable to
assume that fluctuations in DO may not occur immediately in response to
nutrient pulses or freshwater inflow. Work has has shown that many water
quality parameters may have lagged effects lasting days or even months
following storms and large discharge events
\autocite{mooneyWatershedExportEvents2012a,wetzExtremeFutureEstuaries2013,bukaveckasInfluenceStormEvents2020,walkerTimescalesMagnitudeWater2021}.
However, this study only evaluated responses to loading and inflows
occurring the day of water quality observations.

>>>>

\usepackage{csquotes}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\begin{abstract}
Evaluating estuary water quality responses to reductions (or increases)
in nutrient loading attributed to on the ground management actions can
be challenging due to the strong influence of environmental drivers on
nutrient loads and non-linear relationships. This study applied
semiparametric approaches to calculate watershed nutrient loads and
assess responses in estuary water quality to seasonally-adjusted
freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas.
Lavaca Bay is a secondary embayment on the Texas coast displaying early
potential for eutrophication and water quality degradation. Use of
flow-adjusted nutrient loads allowed the study to evaluate the response
in water quality to changes in nutrient loads driven by anthropogenic
sources. Cross-validation indicated that, despite data constraints,
semiparametric models performed well at nutrient load prediction. Based
on these models, delivered annual nutrient loads varied substantially
from year to year. In contrast, minimal changes in flow-normalized loads
indicate that nutrient loadings were driven by natural variation in
precipitation and runoff as opposed to changes in management of nonpoint
sources. Models indicated no evidence of long-term changes in dissolved
oxygen or chlorophyll-\emph{a} within Lavaca Bay. However, site specific
long-term increases in both organic and inorganic nitrogen are
concerning for their potential to fuel eutrophication. Further analysis
found freshwater inflow had strong influences on nutrient and
chlorophyll-\emph{a} concentrations but there was no evidence that
changes in watershed nutrient loading explained additional variation in
dissolved oxygen or chlorophyll-\emph{a} concentrations. In addition to
providing a baseline assessment of watershed nutrient loading and water
quality responses in the Lavaca Bay watershed, this study provides
methodological support for the use of semiparametric models in load
regression models and estuary assessments.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

There are ongoing efforts between local, state, and federal agencies to
address water quality impairments in the freshwater portions of the
Lavaca Bay watershed
\autocite{jainTechnicalSupportDocument2021,schrammLavacaRiverWatershed2018,bertholdDirectMailingEducation2021}.
However, on a statewide scale, these approaches have shown limited
success and emphasize a need for improved methods to assess and link
management actions with downstream water quality
\autocite{schrammTotalMaximumDaily2022}. Such methods could help
identify and replicate effective water quality management actions across
the state . The identification and communication of changes and trends
in water quality is complicated by the fact that trends are often
non-linear and confounded by precipitation and runoff that hinder
traditional analysis
\autocite{wazniakLinkingWaterQuality2007,lloydMethodsDetectingChange2014}.
The development and application of flexible statistical methods such as
Weighted Regressions on Time, Discharge and Season
\autocite[WRTDS,][]{hirschWeightedRegressionsTime2010} and Generalized
Additive Models \autocite[GAMs,][]{woodFastStableRestricted2011} have
provided effective tools for researchers to quantify and communicate
non-linear changes in river and estuary pollutant loadings.

@@

While WRTDS is a statistical approach developed specifically for water
quality applications, GAMs are a broadly applicable statistical method.
GAMs are a semiparametric extension of generalized linear models where
the response variable is modeled as the sum of multiple unknown smooth
functions and parametric linear predictors
\autocite{woodFastStableRestricted2011}. Although the underlying
parameter estimation procedure of GAMs are substantially different than
WRTDS, both the functional form and results have been demonstrated to be
similar when assessing nutrient concentration trends
\autocite{beckNumericalQualitativeContrasts2017}. Water quality
applications of GAMs have included river and catchment nutrient
concentration and load estimation
\autocite{wangLoadEstimationUncertainties2011,kroonRiverLoadsSuspended2012,kuhnertQuantifyingTotalSuspended2012,robson_prediction_2015-1,hagemannEstimatingNutrientOrganic2016,mcdowell_implications_2021,biagi_novel_2022},
and assessment of temporal trends of nutrients
\autocite{beckNumericalQualitativeContrasts2017,murphyGeneralizedAdditiveModel2019},
phytoplankton \autocite{bergbuschUnexpectedShiftPhytoplankton2021}, and
cyanobacteria \autocite{hayesEffectsLakeWarming2020}. Recently GAMs have
also been used to link water quality responses in receiving water bodies
to changes in nonpoint source nutrient inputs
\autocite{murphyNutrientImprovementsChesapeake2022}. For a substantial
discussion on the differences (and similarities) between GAMs and WRTDS
for water quality applications readers are referred to
\textcite{beckNumericalQualitativeContrasts2017}.

@@

Lavaca Bay is a 190 km\textsuperscript{2} estuary with the majority of
freshwater inflow provided by the Lavaca and Navidad River systems
(Fig.~\ref{fig:fig1}). Lavaca Bay is relatively shallow with an average
depth of 1.2 m and a generally well mixed and turbid water column
\autocite{beserespollackLongtermTrendsResponse2011,montagnaAssessmentRelationshipFreshwater2020}.
The Garcitas-Arenosa, Placedo Creek, and Cox Bay subwatersheds provide
additional freshwater inflows. The proximity to freshwater inputs in the
upper Lavaca Bay results in a strong mean salinity and variance gradient
\autocite{montagnaAssessmentRelationshipFreshwater2020}. Average
salinity ranges from 12 psu near the Lavaca River mouth to 23 psu at the
mouth of Lavaca Bay. The variance is inversely correlated with mean
salinity. The entire watershed area is 8,149 km\textsuperscript{2} and
primarily rural. Watershed land cover and land use is 50\% grazed
pasture and rangeland, 20\% cultivated cropland (primarily rows crops
such as corn, cotton, and sorghum), and 5\% suburban/urban. Pasture and
rangeland is concentrated in the Lavaca River watershed, while
cultivated crops are generally located along the eastern tributaries of
the Navidad river. The Lavaca and Navidad River watersheds are a
combined 5,966 km\textsuperscript{2}, or approximately 73\% of the
entire Lavaca Bay watershed area. Discharge from the Navidad River is
regulated by Lake Texana which has been in operation since 1980. Lake
Texana provides 0.210 km\textsuperscript{3} of water storage and
discharges into the tidal section of the Navidad River which ultimately
joins the tidal section of the Lavaca River 15 km upstream of the
confluence with the Lavaca Bay.

@@

{\centering \includegraphics[width=5.2in,]{Schramm-2023-08-PeerJ_files/figure-latex/fig1}

@@

\caption{Map of Lavaca Bay watershed. The freshwater sites are the most downstream freshwater stream locations with water quality and streamflow data used for nutrient load models. Water quality concentration data at the three Lavaca Bay sites were used to assess relationships between freshwater flows, loads and estuary water quality.}\label{fig:fig1}
\end{figure}

Daily discharges for the Lavaca River (USGS-08164000,
Fig.~\ref{fig:fig1}) were obtained from the United States Geologic
Survey (USGS) National Water Information System
(\url{https://waterdata.usgs.gov/nwis}) using the \emph{dataRetrieval} R
package \autocite{deciccoDataRetrievalPackagesDiscovering2022}. Gaged
daily discharges from the outlet of Lake Texana on the Navidad River
(USGS-0816425) were provided by the Texas Water Development Board (TWDB)
(April 21, 2022 email from R. Neupane, TWDB).

@@

\caption{\label{tab:table1}Summary of gauged streamflow and freshwater water quality samples between January 1, 2000 and December 31, 2020.}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}llllrr}
\toprule
Station ID and Watershed & Parameter & Mean & SD & N & Method & AWRL\textsuperscript{a} & Criteria\textsuperscript{b}\\
\midrule
 & TP (mg/L) & 0.21 & 0.09 & 80 & EPA 365.1 & 0.06 & 0.69\\

 & NO\textsubscript{3} (mg/L) & 0.18 & 0.24 & 74 & EPA 353.2 & 0.05 & 1.95\\

\multirow{-3}{2.5cm}{\raggedright\arraybackslash USGS-08164000, Lavaca River} & Mean Daily Streamflow (cfs) & 332.78 & 1667.47 & 7671 &  &  & \\
\cmidrule{1-8}
 & TP (mg/L) & 0.20 & 0.08 & 81 & EPA 365.1 & 0.06 & 0.20\\

 & NO\textsubscript{3} (mg/L) & 0.29 & 0.26 & 62 & EPA 353.2 & 0.05 & 0.37\\

\multirow{-3}{2.5cm}{\raggedright\arraybackslash USGS-08164525, Navidad River} & Mean Daily Streamflow (cfs) & 666.14 & 2957.79 & 7671 &  &  & \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item[a] Ambient water reporting limit (AWRL) represents the highest concentration that can be used as a reporting limit by a laboratory for inclusion in the state's monitoring data.
\item[b] The state of Texas has not adopted numeric criteria for nutrients, these criteria represent screening levels used by the state for assessment purposes.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

\caption{\label{tab:table2}Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.}
\centering
\begin{threeparttable}
\begin{tabular}[t]{lllll>{\raggedright\arraybackslash}p{3cm}rr}
\toprule
Station ID & Parameter & Mean & SD & N & Method & AWRL\textsuperscript{a} & Criteria\textsuperscript{b}\\
\midrule
 & TP (mg/L) & 0.11 & 0.05 & 47 & EPA 365.1 & 0.06 & 0.21\\

 & NO\textsubscript{x} (mg/L) & 0.07 & 0.15 & 51 & EPA 353.2 & 0.05 & \\

 & TKN (mg/L) & 0.94 & 0.49 & 45 & EPA 351.2 & 0.20 & \\

 & Chlorophyll-\textit{a} (ug/L) & 9.43 & 5.31 & 47 & EPA 445.0 & 3.00 & 11.60\\

\multirow{-5}{*}{\raggedright\arraybackslash TCEQ-13383} & DO (mg/L) & 7.22 & 1.35 & 55 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\cmidrule{1-8}
 & TP (mg/L) & 0.08 & 0.03 & 51 & EPA 365.1 & 0.06 & 0.21\\

 & NO\textsubscript{x} (mg/L) & 0.06 & 0.08 & 52 & EPA 353.2 & 0.05 & \\

 & TKN (mg/L) & 0.76 & 0.40 & 48 & EPA 351.2 & 0.20 & \\

 & Chlorophyll-\textit{a} (ug/L) & 8.22 & 6.44 & 46 & EPA 445.0 & 3.00 & 11.60\\

\multirow{-5}{*}{\raggedright\arraybackslash TCEQ-13384} & DO (mg/L) & 7.51 & 1.32 & 54 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\cmidrule{1-8}
 & TP (mg/L) & 0.13 & 0.06 & 50 & EPA 365.1 & 0.06 & 0.21\\

 & NO\textsubscript{x} (mg/L) & 0.09 & 0.13 & 53 & EPA 353.2 & 0.05 & \\

 & TKN (mg/L) & 0.94 & 0.37 & 49 & EPA 351.2 & 0.20 & \\

 & Chlorophyll-\textit{a} (ug/L) & 9.67 & 5.33 & 49 & EPA 445.0 & 3.00 & 11.60\\

\multirow{-5}{*}{\raggedright\arraybackslash TCEQ-13563} & DO (mg/L) & 7.91 & 1.34 & 56 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item[a] Ambient water reporting limit (AWRL) represents the highest concentration that can be used as a reporting limit by a laboratory for inclusion in the state's monitoring data.
\item[b] The state of Texas has not adopted numeric criteria for nutrients, these criteria represent screening levels used by the state for assessment purposes. NO\textsubscript{x} and TKN parameters are not currently screened by the state and do not have an associated screening level.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

Estimates of NO\textsubscript{3} and TP loads at the Lavaca River
(USGS-08164000) and the outlet of Lake Texana on the Navidad River
(USGS-08164525) were developed using GAMs relating nutrient
concentration to river discharge, season, and time. Separate models were
fit at each station for each parameter and used to predict nutrient
concentrations for each day in the study period. GAMs were fit using the
\emph{mgcv} package in R which makes available multiple types of smooth
functions with automatic smoothness selection
\autocite{woodFastStableRestricted2011}. The general form of the model
related NO\textsubscript{3} or TP concentration to a long term trend,
season, streamflow, and two different antecedent discharge terms, shown
in Eq. \ref{eq1}:

@@

The calculation of model terms for the Lake Texana site were modified
because daily loads are not a function of natural stream flow processes
alone, but of dam releases and nutrient concentrations at the discharge
point of the lake. \emph{Q}, \emph{ma}, and \emph{fa} terms were
calculated based on total gaged inflow from the 4 major tributaries to
the lake. Thin-plate regression splines were used for \emph{ddate},
\emph{log(1+Q)}, \emph{fa}, and \emph{ma}. A cyclic cubic regression
spline was used for \emph{yday} to ensure the ends of the spline match
(day 1 and day 366 are expected to match). First order penalties were
applied to the smooths of flow-based variables which penalize departures
from a flat function to help constrain extrapolations for high flow
measurements.

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-08-PeerJ_files/figure-latex/fig2}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-08-PeerJ_files/figure-latex/fig3}

@@

{\centering \includegraphics[width=0.5\linewidth,]{Schramm-2023-08-PeerJ_files/figure-latex/fig4}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-08-PeerJ_files/figure-latex/fig5}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-08-PeerJ_files/figure-latex/fig6}

@@

{\centering \includegraphics[width=1\linewidth,]{Schramm-2023-08-PeerJ_files/figure-latex/fig7}

@@

Some prior studies have generated estimates of mean annual TP yields in
the Lavaca River watershed
\autocites[Table~\ref{tab:table5},][]{dunnTrendsNutrientInflows1996,rebichSourcesDeliveryNutrients2011,omaniEstimationSedimentNutrient2014,wise_spatially_2019}.
Although these studies differ in time periods and methodologies, they
provide a sanity check for the reasonableness of the annual estimates
generated in the current study. In a regional assessment of nutrient
loading in river's along the Gulf of Mexico,
\textcite{dunnTrendsNutrientInflows1996} used the LOADEST model to
develop an estimated mean annual yield of 28.9 kg/km\textsuperscript{2}.
LOADEST is a multiple linear regression model that fits log transformed
pollutant concentrations to long term, seasonal, and flow based
predictors and includes methods for bias correction when exponentiating
the response variable. \textcite{rebichSourcesDeliveryNutrients2011} and
\textcite{wise_spatially_2019} used SPARROW to provide a more recent
assessment of regional catchment based loadings to the Gulf of Mexico
(Table~\ref{tab:table5}). SPARROW is a hybrid statistical-process model
with the underlying nutrient load estimation methods based on the
previously described LOADEST
\autocite{schwarzSPARROWSurfaceWaterQuality2006}. The functional form of
the LOADEST regression model is similar to the terms applied in the GAMs
used in the current study. The only study to apply a mechanistic
watershed model (SWAT) to estimate nutrient loadings in the Lavaca River
watershed \textcite{omaniEstimationSedimentNutrient2014} developed
estimated yields (42 kg/km\textsuperscript{2}) similar to the two
SPARROW models. Although direct comparisons are complicated by varying
time periods, the estimates in this study do fall within the range of of
previously developed estimates. To evaluate changes in long-term trends
in discharge might be associated with the nutrient yield estimates
covering different time periods, we fit a GAM relating log-transformed
daily discharges on the Lavaca River to season and time
(Fig.~\ref{fig:fig8}). The long-term trends in discharge indicate
watershed discharges were at or above average from 1972 though the early
and mid-1980s. In comparison watershed discharges since the mid-2000s
are at or below average. It is probable that the lower than average
discharges observed from 2010 through 2021 (Fig.~\ref{fig:fig8}) bias
our estimates downward compared to studies that included higher than
average streamflow periods (1995-2005). Overall, the ranges of estimated
yields among different studies along with the apparent large variability
in streamflow driven loadings (Fig.~\ref{fig:fig3}, Fig. \ref{fig:fig4})
suggest that the current estimates of TP loading are reasonable.

@@

{\centering \includegraphics[width=0.7\linewidth,]{Schramm-2023-08-PeerJ_files/figure-latex/fig8}

@@

\caption{Measured daily discharges (log-transformed) and smoothed long-term trends for the Lavaca River form 1972 though 2001.}\label{fig:fig8}
\end{figure}

@@

There was no evidence that adjusted freshwater inflow and nutrient loads
had effects on DO concentration in Lavaca Bay. The seasonality term in
the temporal GAM models explained a substantial amount of DO variation
at all of the sites. Responses of estuary metabolic processes and
resulting DO concentrations can be quite complicated and often locally
specific \autocite{caffreyFactorsControllingNet2004}. While the lack of
total nitrogen or TKN loading data hinders interpretation, the large
seasonal effect on DO concentration indicates physical factors (such as
temperature, wind, and turbidity) play an important role and should be
included in future models. Prior work suggests that Lavaca Bay may not
be limited by nutrients alone, with high turbidity or nutrient
processing in upper portions of the Bay or intertidal river limiting
production \autocite{russell_effect_2006}. Finally, it is reasonable to
assume that fluctuations in DO may not occur immediately in response to
nutrient pulses or freshwater inflow. Work has shown that many water
quality parameters may have lagged effects lasting days or even months
following storms and large discharge events
\autocite{mooneyWatershedExportEvents2012a,wetzExtremeFutureEstuaries2013,bukaveckasInfluenceStormEvents2020,walkerTimescalesMagnitudeWater2021}.
However, this study only evaluated responses to loading and inflows
occurring the day of water quality observations.
